Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-02-10 WSSSPE2_report/WSSSPE2_review.tex
adding some text for understanding sustainability section

\todo{Dan to discuss initial breakout and results}

>>>>

\todo{writing led by Dan - other contributions welcome}

In the first interactive session, the attendees divided themselves into groups to discuss software sustainability.  They were asked to
\begin{enumerate}
\item discuss what the term ``software sustainability'' meant to them
\item determine three things they considered to be significant enablers of software sustainability
\item determine three things they considered to be significant barriers to software sustainability
\end{enumerate}
Once each group had come up with answers, all the answers were compiled, and the attendees voted on which they thought were important by a show of hands.

The general responses to what software sustainability meant were:

\begin{itemize}
\item keeping software scientifically useful
\item separating techniques in code from knowledge in code
\item that an adequately large community finds value in software and is willing to sustain it
\end{itemize}

\katznote{any discussion of this?}

The enablers of and barriers to software sustainability, roughly ranked by attendee votes are shown in Tables~\ref{tb:software_sustainability_enablers} and \ref{tb:software_sustainability_barriers}, respectively.\footnote{A few other items were suggested as barriers, but were not voted on due to lack of time in the session:
layering up dependencies;
using software past its sustainable life;
using software past its usable life;
inertia for accepted answers vs wrong/right answers;
monolithic or poor code; and
need to restructure code when hardware/software/libraries change.
}

\katznote{any discussion of these tables?}

\begin{table*}[ht]
\begin{center}
    \caption{Enablers of software sustainability, with 0 to 10 `*'s roughly indicating the fraction of attendees who voted for an item as important}
    \begin{scriptsize}
    \begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
    \hline
  Importance & item \\ \hline \hline
********** & healthy and vibrant communities; vibrant community to champion software \\ \hline
********** & designing for growth and extension - open development \\ \hline
******* & culture in community for reuse \\ \hline
**** & portability \\ \hline
**** & culture in dev community to support transition between developers \\ \hline
*** & interdisciplinary people : science + IT experience \\ \hline
** & planning for end of life \\ \hline
** & make smart choices about dependencies \\ \hline
* & thinking of software as product lines - long term vs short term view \\ \hline
 & not all communities need new software \\ \hline
 & converting use into resources \\ \hline
    \end{tabular}
    \end{scriptsize}
    \label{tb:software_sustainability_enablers}
\end{center}   
\end{table*}

\begin{table*}[ht]
\begin{center}
    \caption{Barriers to software sustainability, with 0 to 10 `*'s roughly indicating the fraction of attendees who voted for an item as important}
    \begin{scriptsize}
    \begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
    \hline
  Importance & item \\ \hline \hline
******* & lack of incentives, including promotion and tenure process; promotion and tenure process in academic is incompatible \\ \hline
*****  & absent / poor documentation \\ \hline
***** & funding difficult \\ \hline
**** & developers are not computer scientists; don't have software engineering practices (in particular, those needed to scale-up projects to support and be developed by a large sustainable community) \\ \hline
*** & overreliance on one or two people - bus test \\ \hline
** & rate of change of underlying technologies \\ \hline
** & lack of business models for sustainability \\ \hline
* & lack of training for how to build sustainability into the system \\ \hline
 & maintenance needed for software is not visible, appears to ``just happen'' \\ \hline
 & licensing issues \\ \hline
 & staff turnover - lack of continuity \\ \hline
    \end{tabular}
    \end{scriptsize}
    \label{tb:software_sustainability_barriers}
\end{center}   
\end{table*}
